Copulas and their potential for ecology: Appendices

### 2019-12-20 SupportingInformation.Rmd
suppmat checked, need to fig out table s1

---
title: "Copulas and their potential for ecology: Appendices for main text"
author: "Shyamolina Ghosh, Lawrence W. Sheppard, Mark T. Holder, Terrance D. Loecke, Philip C. Reid, James D. Bever, Daniel C. Reuman"
geometry: "left=1cm,right=1cm,top=2.2cm,bottom=2.2cm"

@@

The data of Fig. \ref{MT-fig_cop_pedag1}A-B were generated to have standard normal marginals.
Fig. \ref{fig_transf_pearson_spearman}A can be obtained from Fig. \ref{MT-fig_cop_pedag1}A via transformation
of both variables by the cummulative distribution function (cdf) of the standard normal distribution, and then by the inverse of 
the cdf of the gamma distribution with shape and scale parameters 2. 
The functions `pnorm` and `qgamma` in R were used.
Fig. \ref{fig_transf_pearson_spearman}B can be obtained from Fig. \ref{MT-fig_cop_pedag1}B in the same way.
The Pearson correlations differ substantially after transformation, 
but the Spearman correlations are unchanged (Figs \ref{MT-fig_cop_pedag1}A-B, \ref{fig_transf_pearson_spearman}).
A scatterplot very simlar to Fig. \ref{MT-fig_cop_pedag1}C can be obtained from Fig. \ref{MT-fig_cop_pedag1}A via transformation
of both variables by the cdf of the standard normal distribution.
Likewise a scatterplot similar to Fig. \ref{MT-fig_cop_pedag1}D can be obtained from Fig. \ref{MT-fig_cop_pedag1}B in the same way.

@@

\textbf{Environmental data, soil C and N.} Measurments were from the central pedon of a soil 
core, to 1m depth.

@@

knitr::kable(NA, "latex", booktabs = T, linesep = "", escape=F,
             caption.short = "Summary of principal notation",
      caption = "Summary of principal notation. Many notations are defined and used only within a small part of the paper. This table summarizes notation that is used across larger portions of the paper. Section headers refer to the section in which the notation was originally defined. cdf = cumulative distribution function. \\label{tab_summary_notation}")
```

@@

\noindent The matrices from which root character states and changes 
across phylogeny branches were randomly chosen were generated as follows.
For extreme right-tail dependent noise, we first produced
a 1,000,000 by $2$ matrix, $m$, filled with 2,000,000 
independent standard-normal random draws. Then, each row of $m$ 
was identified for a first set of modifications randomly 
and independently with a $50\%$ chance; 
for those rows identified, every element of the row was replaced with the 
absolute value of the first element of the row (including the first element itself).
For other rows, every element in the row was replaced by the negative of its
own absolute value. This produces right-tail dependent noise, normally distributed
for each of the two locations. 
If left-tail dependent noise was desired, we took the negative of the
entire matrix. The moderately left-tail dependent matrix was based on a Clayton copula 
with parameter chosen (using the `iRho` function in the `copula` package) so that
the Spearman correlation would be the same as for the extreme tail dependence 
datasets. The moderately right-tail dependent matrix and the symmetric tail dependence
matrix were constructed similarly, but using an survival Clayton copula and an normal copula,
respectively.

@@

- Given data $x_i(t)$ for $i=1,\ldots,N$ and $t=1,\ldots,T$, let $\tau_{ij}$ be the Spearman (or Kendall) 
correlation of $x_i(t)$ and $x_j(t)$.
- Calculate the covariance matrix of a multivariate normal distribution with standard-normal
marginals and with pairwise Spearman (or Kendall) correlations equal to the $\tau_{ij}$. This is possible.
using the `iRho` function (or the `iTau` function) of the `copula` package
- Generate data $a_i(t)$ for $i=1,\ldots,N$ and $t=1,\ldots,T$ by taking $T$ independent draws
from the multivariate normal distribution of the previous step.
- For any $i,t$ for which the datum $x_i(t)$ was missing, replace $a_i(t)$ by `NA`.
- For each $k$ and $i$ for which $a_i(t)$ is not `NA`, replace the 
$k$th-smallest (non-`NA`) element of the time series 
$a_i(t)$, $t=1,\ldots,T$ by the $k$th smallest (non-`NA`) 
element of $x_i(t)$, $t=1,\ldots,T$.
Call the result $b_i(t)$. This is a surrogate dataset.

@@

We then generated a series of arrays that were the same dimensions ($N \times n \times 1000$)
as the normal-copula population array, and that were statistically similar to 
the normal-copula population array in all ways except the new
arrays were constructed to have Clayton (respectively, Frank, survival Clayton) copula structure for 
the dependences between 
populations in different sampling locations within a simulation. We 
refer to the new arrays as our
*Clayton-copula*, *Frank-copula*, and *Survival Clayton-copula population arrays*, respectively. The Clayton-copula array was
constructed as follows; the others were constructed analogously. First, we constructed an
$n$-dimensional Clayton copula with parameter chosen so that the 
Spearman correlations between each pair of the $n$ components were the same as those
of the normal distribution constructed above. This was done using the `rho` and `iRho` functions 
in the `copula` package in R. Because the Clayton copula is a one-paremeter family, this 
step relies on the fact that all pairs of locations were 
equally correlated with each other in the original multivariate normal distribution
constructed above. We then randomized each simulation in 
the normal-copula population array in a special way 
(see Appendix \ref{sect:copsurrognd}) so that the resulting 
surrogate/randomized dataset had the copula structure
of the Clayton copula just constructed. The randomization procedure worked 
by permuting time series of values in the normal-copula array, separately for each 
sampling location and simulation. So population marginal distrbutions for each location
and simulation were *exactly* the same as those in the normal-copula population array. 
Permutations were done in such a way
that resulting pairwise Spearman correlations were the same as the Clayton copula 
constructed above, and hence were the same as the normal-copula array, 
except for sampling variation. Thus, to summarize using mathematical notation,
if $m^{(\text{N},k)}_{t,j}$ is our normal-copula population array (the three dimensions 
of the array are indexed, respectively, by $t$, $j$ and $k$, and $m^{(\text{C},k)}_{t,j}$ is our 
Clayton-copula population array, then,
for any given $j$ and $k$, the sets $\{m^{(\text{N},k)}_{t,j}: t=1,\ldots,N\}$ and 
$\{m^{(\text{C},k)}_{t,j}: t=1,\ldots,N\}$ are exactly the same (when considered 
as unordered sets); 
and for any given $j_1$, $j_2$ and $k$ with $j_1 \neq j_2$, the Spearman correlations $\cor_t(m^{(\text{C},k)}_{t,j_1},m^{(\text{C},k)}_{t,j_2})$
and $\cor_t(m^{(\text{N},k)}_{t,j_1},m^{(\text{N},k)}_{t,j_2})$ are the same, to within 
sampling variation. The whole exercise was repeated preserving Kendall instead of 
Spearman correlations.

@@

- Generate data $a_i(t)$ for $i=1,\ldots,N$ and $t=1,\ldots,T$ by taking $T$ independent 
draws from $C$.
- For each $k$ and $i$, replace the $k$th-smallest element of the time series $a_i(t)$, 
$t=1,\ldots,T$ by the $k$th smallest element of $x_i(t)$, $t=1,\ldots,T$. Call the 
result $y_i(t)$. This is the surrogate dataset.

>>>>

---
title: "Copulas and their potential for ecology: Appendices"
author: "Shyamolina Ghosh, Lawrence W. Sheppard, Mark T. Holder, Terrance D. Loecke, Philip C. Reid, James D. Bever, Daniel C. Reuman"
geometry: "left=1cm,right=1cm,top=2.2cm,bottom=2.2cm"

@@

The data of Fig. \ref{MT-fig_cop_pedag1}A-B were generated to have standard normal marginals.
Fig. \ref{fig_transf_pearson_spearman}A can be obtained from Fig. \ref{MT-fig_cop_pedag1}A via transformation
of both variables by the cumulative distribution function (cdf) of the standard normal distribution, and then by the inverse of 
the cdf of the gamma distribution with shape and scale parameters 2. 
The functions `pnorm` and `qgamma` in R were used.
Fig. \ref{fig_transf_pearson_spearman}B can be obtained from Fig. \ref{MT-fig_cop_pedag1}B in the same way.
The Pearson correlations differ substantially after transformation, 
but the Spearman correlations are unchanged (Figs \ref{MT-fig_cop_pedag1}A-B, \ref{fig_transf_pearson_spearman}).
A scatterplot very similar to Fig. \ref{MT-fig_cop_pedag1}C can be obtained from Fig. \ref{MT-fig_cop_pedag1}A via transformation
of both variables by the cdf of the standard normal distribution.
Likewise a scatterplot similar to Fig. \ref{MT-fig_cop_pedag1}D can be obtained from Fig. \ref{MT-fig_cop_pedag1}B in the same way.

@@

\textbf{Environmental data, soil C and N.} Measurements were from the central pedon of a soil 
core, to 1m depth.

@@

knitr::kable(NA, "latex", booktabs = T, linesep = "", escape=F,
             caption.short = "Summary of principal notation",
      caption = "Summary of principal notation. Many notations are defined and used only within a small part of the paper. This table summarizes notation that is used across larger portions of the paper. Third coloumns to the section in which the notation was originally defined. cdf = cumulative distribution function. \\label{tab_summary_notation}")
```

@@

\noindent The matrices from which root character states and changes 
across phylogeny branches were randomly chosen were generated as follows.
For extreme right-tail dependent noise, we first produced
a 1,000,000 by $2$ matrix, $m$, filled with 2,000,000 
independent standard-normal random draws. Then, each row of $m$ 
was identified for a first set of modifications randomly 
and independently with a $50\%$ chance; 
for those rows identified, every element of the row was replaced with the 
absolute value of the first element of the row (including the first element itself).
For other rows, every element in the row was replaced by the negative of its
own absolute value. This produces right-tail dependent noise, normally distributed
for each of the two locations. 
If left-tail dependent noise was desired, we took the negative of the
entire matrix. The moderately left-tail dependent matrix was based on a Clayton copula 
with parameter chosen (using the `iRho` function in the `copula` package) so that
the Spearman correlation would be the same as for the extreme tail dependence 
datasets. The moderately right-tail dependent matrix and the symmetric tail dependence
matrix were constructed similarly, but using a survival Clayton copula and a normal copula,
respectively.

@@

- Given data $x_i(t)$ for $i=1,\ldots,N$ and $t=1,\ldots,T$, let $\tau_{ij}$ be the Spearman (or Kendall) 
correlation of $x_i(t)$ and $x_j(t)$.
- Calculate the covariance matrix of a multivariate normal distribution with standard-normal
marginals and with pairwise Spearman (or Kendall) correlations equal to the $\tau_{ij}$. This is possible.
using the `iRho` function (or the `iTau` function) of the `copula` package
- Generate data $a_i(t)$ for $i=1,\ldots,N$ and $t=1,\ldots,T$ by taking $T$ independent draws
from the multivariate normal distribution of the previous step.
- For any $i,t$ for which the datum $x_i(t)$ was missing, replace $a_i(t)$ by `NA`.
- For each $k$ and $i$ for which $a_i(t)$ is not `NA`, replace the 
$k$th-smallest (non-`NA`) element of the time series 
$a_i(t)$, $t=1,\ldots,T$ by the $k^{th}$ smallest (non-`NA`) 
element of $x_i(t)$, $t=1,\ldots,T$.
Call the result $b_i(t)$. This is a surrogate dataset.

@@

We then generated a series of arrays that were the same dimensions ($N \times n \times 1000$)
as the normal-copula population array, and that were statistically similar to 
the normal-copula population array in all ways except the new
arrays were constructed to have Clayton (respectively, Frank, survival Clayton) copula structure for 
the dependences between 
populations in different sampling locations within a simulation. We 
refer to the new arrays as our
*Clayton-copula*, *Frank-copula*, and *Survival Clayton-copula population arrays*, respectively. The Clayton-copula array was
constructed as follows; the others were constructed analogously. First, we constructed an
$n$-dimensional Clayton copula with parameter chosen so that the 
Spearman correlations between each pair of the $n$ components were the same as those
of the normal distribution constructed above. This was done using the `rho` and `iRho` functions 
in the `copula` package in R. Because the Clayton copula is a one-paremeter family, this 
step relies on the fact that all pairs of locations were 
equally correlated with each other in the original multivariate normal distribution
constructed above. We then randomized each simulation in 
the normal-copula population array in a special way 
(see Appendix \ref{sect:copsurrognd}) so that the resulting 
surrogate/randomized dataset had the copula structure
of the Clayton copula just constructed. The randomization procedure worked 
by permuting time series of values in the normal-copula array, separately for each 
sampling location and simulation. So population marginal distributions for each location
and simulation were *exactly* the same as those in the normal-copula population array. 
Permutations were done in such a way
that resulting pairwise Spearman correlations were the same as the Clayton copula 
constructed above, and hence were the same as the normal-copula array, 
except for sampling variation. Thus, to summarize using mathematical notation,
if $m^{(\text{N},k)}_{t,j}$ is our normal-copula population array (the three dimensions 
of the array are indexed, respectively, by $t$, $j$ and $k$, and $m^{(\text{C},k)}_{t,j}$ is our 
Clayton-copula population array, then,
for any given $j$ and $k$, the sets $\{m^{(\text{N},k)}_{t,j}: t=1,\ldots,N\}$ and 
$\{m^{(\text{C},k)}_{t,j}: t=1,\ldots,N\}$ are exactly the same (when considered 
as unordered sets); 
and for any given $j_1$, $j_2$ and $k$ with $j_1 \neq j_2$, the Spearman correlations $\cor_t(m^{(\text{C},k)}_{t,j_1},m^{(\text{C},k)}_{t,j_2})$
and $\cor_t(m^{(\text{N},k)}_{t,j_1},m^{(\text{N},k)}_{t,j_2})$ are the same, to within 
sampling variation. The whole exercise was repeated preserving Kendall instead of 
Spearman correlations.

@@

- Generate data $a_i(t)$ for $i=1,\ldots,N$ and $t=1,\ldots,T$ by taking $T$ independent 
draws from $C$.
- For each $k$ and $i$, replace the $k^{th}$-smallest element of the time series $a_i(t)$, 
$t=1,\ldots,T$ by the $k^{th}$ smallest element of $x_i(t)$, $t=1,\ldots,T$. Call the 
result $y_i(t)$. This is the surrogate dataset.
